A highly-available move operation for replicated trees

### 2019-06-14 paper/move-op.tex

As the tree structure may be concurrently modified on different replicas, the state of these replicas may temporarily diverge.
In this paper we show how the replicas can nevertheless achieve \emph{strong eventual consistency}~\cite{Shapiro:2011un}: as the replicas communicate, we guarantee that they will converge towards a consistent state.
Our algorithm for achieving convergence is an example of a \emph{Conflict-free Replicated Data Type} or CRDT~\cite{Shapiro:2011wy,Shapiro:2011un}.

@@

@@

% Related work:
% OT implementation of move operation: https://hal.inria.fr/inria-00107740/
% Ficus (does not discuss move): https://pdos.csail.mit.edu/6.824/papers/ficus.pdf
% Roam: https://ieeexplore.ieee.org/abstract/document/795151

>>>>

As the tree structure may be concurrently modified on different replicas, the state of these replicas may temporarily diverge.
In this paper we show how the replicas can nevertheless achieve \emph{strong eventual consistency}~\cite{Shapiro:2011un,Gomes:2017gy}: as the replicas communicate, we guarantee that they will converge towards a consistent state.
Our algorithm for achieving convergence is an example of a \emph{Conflict-free Replicated Data Type} or CRDT~\cite{Shapiro:2011wy,Shapiro:2011un}.

@@

TODO

@@

Many replicated data systems use optimistic replication~\cite{Saito:2005jw}, which allows the state of replicas to temporarily diverge, in order to achieve better performance and better availability in the presence of faults than strongly consistent systems~\cite{Bailis:2014th,Gilbert:2002il}.
As a consequence, these systems require a mechanism for merging or reconciling conflicting updates that were made concurrently on different replicas.
For example, version control systems such as Git~\cite{Chacon:2014kr} leave any conflicts for the user to resolve manually.
Some systems, such as Dynamo~\cite{DeCandia:2007ui} and Bayou~\cite{Terry:1995dn}, rely on the application programmer to provide explicit conflict resolution logic; however, such logic is difficult to get right \cite{Bailis:2013jc,Burckhardt:2014hy,Gomes:2017gy}.
Hence, we focus on systems that automatically ensure that all replicas converge towards a consistent state, without requiring custom application logic---a consistency model known as \emph{strong eventual consistency}~\cite{Shapiro:2011un,Gomes:2017gy}.

\subsection{Conflict-free Replicated Data Types}

Our algorithm is an example of an operation-based Conflict-free Replicated Data Type or CRDT~\cite{Shapiro:2011wy,Shapiro:2011un,Burckhardt:2014ft}.
All CRDTs share the property that concurrent changes on different replicas can be merged in any order; any two replicas that have seen the same set of updates are guaranteed to be in the same state, regardless of the order in which they processed these updates.
CRDTs have been defined for various abstract datatypes, such as registers and counters \cite{Shapiro:2011wy,Shapiro:2011un}, maps \cite{Baquero:2016iv,Kleppmann:2016ve}, sets \cite{Bieniusa:2012wu,Bieniusa:2012gt}, lists (sequences) \cite{Roh:2011dw,Nedelec:2013ky}, and text \cite{Preguica:2009fz,Weiss:2010hx}.
For tree data structures, several approaches have been proposed:
\begin{itemize}
    \item Martin et al.~\cite{Martin:2010ih} define a CRDT for XML data, and Kleppmann and Beresford~\cite{Kleppmann:2016ve} define a CRDT for JSON.
        However, these algorithms only deal with insertion and deletion of tree nodes, and do not support a move operation.
    \item As discussed in Section~\ref{sec:impossibility}, Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} propose two alternative implementations for a replicated filesystem: a CRDT in which conflicting moves are handled by duplicating tree nodes (as in Figures~\ref{fig:move-same-item}b and~\ref{fig:move-cycle}b), and a centralised implementation in which move operations must obtain a lock before executing (not a CRDT since it relies on synchronous coordination).
    \item Ahmed-Nacer et al.~\cite{AhmedNacer:2012us} survey approaches to handling conflicts on trees without providing concrete algorithms.
    \item Tao et al.~\cite{Tao:2015gd} propose handling conflicting move operations by allowing the same object to appear in more than one location; thus, their datatype is strictly a DAG, not a tree.
        Some conflicts are handled by duplicating tree nodes.
    \item In our previous unpublished work \cite{ExtendedVersion} we develop the foundations for the algorithm described in this paper.
\end{itemize}

Besides CRDTs, another family of algorithms for concurrent modification of data structures is called \emph{Operational Transformation}~\cite{Sun:1998vf}.
Several authors have defined collaboratively editable tree structures using the Operational Transformation approach \cite{Jungnickel:2016cb,Ignat:2003jy,Davis:2002iv}, but they only consider insertion and deletion of tree nodes, and do not support a move operation.

Molli et al.~\cite{Molli:2003cd} define an Operational Transformation tree structure with a move operation.
However, it requires that all communication between replicas is performed via total order broadcast, which in turn requires the use of a sequencer server or a consensus algorithm~\cite{Chandra:1996cp}.
Our algorithm has better availability characteristics in the presence of network partitions because it allows messages to be delivered in any order.

\subsection{Distributed filesystems}

TODO: Coda~\cite{Kumar:1995wf}, Ficus~\cite{Reiher:1994wh}, Rumor~\cite{Guy:1999gy}, Roam~\cite{Ratner:1999fh}

\section{Conclusions}

TODO
